*Supplement* Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-02-08 manu-draft.qmd
redo figs, tables with new gams
more CA edits

Seagrasses have long been a focal point for management efforts aimed at restoring ecosystem health in estuaries worldwide. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay during the summer months. Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at fixed stations in the major bay segments since the early 1970s (Figure [-@fig-map]b). The 45 stations with the longest and most complete temporal record from 1975 to present were used herein. Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Most analyses herein used only bottom water measurements given the shallow depth and mixed water column of most of Tampa Bay [@weisberg2006]. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. Compared to the additional datasets described below, the monitoring stations are generally in deeper water beyond where seagrasses occur along the shallow margins of the bay. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

@@

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, Tables [-@tbl-temptrndtab], [-@tbl-saltrndtab], Figures S1, S2, S3). Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends. The strongest trends were observed for the EPC dataset. The top and bottom water temperature or salinity changes were similar across bay segments (@fig-meteowqraw). Trends in water temperature were similar across bay segments with significant increases for all for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.4 (OTB) to 2.0 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab). Salinity trends were also similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.7 (OTB) and 2.1 (LTB) ppt from 1975 to present.

@@

The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022). All stations had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a). Seasonally, most bay segments had more significant stations during the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed throughout the bay. Temperature trends were more often observed in the summer or early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends. Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.

@@

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB. Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022). Trends for the other two temperature thresholds were similar.

The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural salinity gradient along the bay's longitudinal axis (Tables S3, S4). None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt. The only significant model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.7 days per year. Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB. The number of days per year when salinity was below 25 ppt in OTB increased by 2 days per year, whereas the number of days in HB increased by 0.9 days per year from 1975 to present. Likewise, the mean number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).

The number of days when both temperature was above and salinity was below a threshold also varied by bay segment (Table S3). The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively. Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period of record varied from 6, 12, and 1 day(s) in 1975 to 72, 68, and 34 days in 2022 for OTB, HB, and MTB, respectively.

@@

Global increases in temperature and altered precipitation patterns related to climate change have had measurable affects on the structure and functioning of a wide range of natural environments [@osland2015; @oliver2018]. For Tampa Bay, these changes have been demonstrated using long-term trends in water temperature and salinity, which mirrored long-term changes in air temperature and precipitation. Over the previous fifty years, Tampa Bay has gotten hotter and fresher; water temperature has increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most cearly observed in the dataset with the longest period of record (EPC), covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in known seagrass stressors; the continuous number of days increased each year when temperature, salinity, or both crossed thresholds.

Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@carlson2018; @nickerson2023; @shiir]. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico. Temperature trends were similar to those herein for Tampa Bay (the EPC dataset was also used). @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, although they rightfully acknowledge the sensitivity of their results to conditions at the start and end of the time series. Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain. Our results showing increases in temperature and decreases in salinity for OTB and HB for the summer months are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico. These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths. Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico. The upper limit of our warming estimate for Tampa Bay is comparable. Notably, @carlson2018 suggest a link between historical seagrass losses in Florida Bay and rapid warming in shallow areas with low surface reflectance.

>>>>

Seagrasses have long been a focal point for management efforts aimed at restoring ecosystem health in estuaries worldwide. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay. Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by climate change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at fixed stations in the major bay segments since the early 1970s (Figure [-@fig-map]b). The 45 stations with the longest and most complete temporal record from 1975 to present were used herein. Water quality samples are collected at each station from surface water grabs (e.g., nutrients, biological, and chemical constituents) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Most analyses herein used only bottom water measurements given the shallow depth and mixed water column of most of Tampa Bay [@weisberg2006], although 1975 bottom salinity used middle water column sampling since the former was not available until the following year. Most samples are collected from mid-morning to early afternoon. Compared to the additional datasets described below, the monitoring stations are generally in deeper water beyond where seagrasses occur along the shallow margins of the bay. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

@@

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, Tables [-@tbl-temptrndtab], [-@tbl-saltrndtab], Figures S1, S2, S3). Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends. The strongest trends were observed for the EPC dataset. The top and bottom water temperature or salinity changes were similar across bay segments (@fig-meteowqraw). Trends in water temperature were similar across bay segments with significant increases for all for all four bay segments varying from 0.03 to 0.04 $^\circ$C per year, with a total change from 1.3 (OTB) to 1.7 (HB) $^\circ$C across the period of record from 1975 to 2022 (@tbl-temptrndtab). Salinity trends were also similar between bay segments, although overall salinity was predictably higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.6 (OTB) and 1.6 (LTB) ppt from 1975 to present.

@@

The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022). All stations had significantly increasing temperature and decreasing salinity from 1975 to present, although several stations in HB did not have significant salinity trends, the exception being the northernmost stations near the mouth of the Hillsborough Rive (Figure [-@fig-kendall]a). Seasonally, most bay segments had more significant stations during the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed throughout the bay. Temperature trends were more often observed in the summer or early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends. Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.

@@

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.5 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from less than 1 day per year in LTB to 1.1 days per year in MTB. Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied between bay segments, although most had a general increase of 50 days each year across the bay segments for the period of record (Table S5, see Table S6 for 1998 to 2022). The increase in the number days each year when temperatures were above 29 or 31 $^\circ$C was similar.

The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural salinity gradient along the bay's longitudinal axis (Tables S3, S4). None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt. The only significantly increasing model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.5 days per year. Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were more than twice those in HB. Some salinity models had significantly decreasing trends, although the total number of days at the beginning and end of the period of record were neglible (Table S5). Overall, OTB showed the largest increase in the number of days each year when salinity was below a theshold, particularly for 25 ppt, where the change was 86 days per year from 1975 to 2022 (130 days to 216 days per year, Table S5, see Table S6 for 1998 to 2022).

The number of days when both temperature was above and salinity was below a threshold also varied by bay segment (Table S3). The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.1, and 0.7 days per year for OTB, HB, and MTB, respectively. Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (4 days to 72 days per year from 1975 to 2022).

@@

Global increases in temperature and altered precipitation patterns related to climate change have had measurable affects on the structure and functioning of a wide range of natural environments [@osland2015; @oliver2018]. For Tampa Bay, these changes have been demonstrated using long-term trends in water temperature and salinity, which mirrored long-term changes in air temperature and precipitation. Over the previous fifty years, Tampa Bay has gotten hotter and fresher; water temperature has increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most clearly observed in the dataset with the longest period of record (EPC), covering nearly fifty years of monthly observations. These long-term changes manifested into consistent trends in known seagrass stressors; the continuous number of days increased each year when temperature, salinity, or both crossed thresholds.

Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@carlson2018; @nickerson2023; @shiir]. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico. Temperature trends were similar to those herein for Tampa Bay (the EPC dataset was also used). @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, although they rightfully acknowledge the sensitivity of their results to conditions at the start and end of the time series. Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain. Our results showing increases in temperature and decreases in salinity in the summer months, most notably for OTB and northern stations of HB [@fig-kendall], are likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico. These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths. Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico. The upper limit of our warming estimate for Tampa Bay is comparable. Notably, @carlson2018 suggest a link between historical seagrass losses in Florida Bay and rapid warming in shallow areas with low surface reflectance.

### 2024-02-08 supplement.qmd
shorten text, remove hydro load from figs
more CA edits

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "*Supplement* Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      institute: [TBEP]
      correspondence: false
  - Sheila Scolaro:
      email: sscolaro@tbep.org
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      institute: [TBEP]
      correspondence: false
  - Edward T Sherwood:
      email: esherwood@tbep.org
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA

@@

![Air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022. The color shades for water temperature and salinity indicate sampling location and values shown are the averages (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}

>>>>

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "*Supplement* Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      institute: [TBEP]
      correspondence: false
  - Sheila Scolaro:
      email: sscolaro@tbep.org
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      institute: [TBEP]
      correspondence: false
  - Bradley T Furman:
      email: Bradley.Furman@myfwc.com
      institute: [FWRI]
      correspondence: false
  - David J Karlen: 
      email: karlen@epchc.org
      institute: [EPCHC]
      correspondence: false
  - Christopher Pratt: 
      email: prattc@epchc.org
      institute: [EPCHC]
      correspondence: false
  - Christopher J Anastasiou: 
      email: chris.anastasiou@swfwmd.state.fl.us
      institute: [SWFWMD]
      correspondence: false    
  - Edward T Sherwood:
      email: esherwood@tbep.org
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA
  - FWRI: Florida Fish and Wildlife Conservation Commission, Florida Fish and Wildlife Research Institute, 100 8th Avenue SE, St. Petersburg, FL 33701, USA
  - EPCHC: Environmental Protection Commission of Hillsborough County, 3629 Queen Palm Drive, Tampa, FL 33619, USA
  - SWFWMD: Southwest Florida Water Management District, 7601 US Highway 301 North, Tampa, FL 33637, USA

@@

![Air temperature, precipitation (Jun-Aug), Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022. The color shades for water temperature and salinity indicate sampling location and values shown are the averages (95% confidence interval) across all Environmental Protection Commission (EPC) stations in each bay segment and sampling months for each year. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}
